Constraints on Cosmic-Ray Particle Populations, Magnetic Fields, and Dark Matter from VHE Gamma-Ray Observations of the Coma Cluster of Galaxies with VERITAS

### 2011-07-28 ms.tex
Added two tables for the DM results: astrophysical factors and <sigma v> limits

A third mechanism for gamma-ray production in a galaxy cluster is self-annihilation of a DM particle, e.g. a weakly interacting massive particle (WIMP). As already mentioned, about 80\% of the cluster mass is in the form of dark matter. Recent works \citep{article:EvansFerrerSarkar:2004, article:BergstromHooper:2006} have shown that the gamma-ray emissivity from DM annihilations is expected to be proportional to the DM density along the line of sight, making clusters of galaxies potentially good candidates for DM searches.

@@

\begin{deluxetable}{lcc}
\tablecolumns{3}
\tablewidth{0pc}
\tablecaption{Regions of interest in the Coma cluster field of view.}
\tablehead{
\colhead{Source} &
\colhead{RA} &
\colhead{Dec}
}
\startdata
Core & \RA{12}{59}{48.7} & \Dec{+27}{58}{50.0}\\
NGC 4889 & \RA{13}{00}{08.13} & \Dec{+27}{58}{37.03}\\
NGC 4874 & \RA{12}{59}{35.71} & \Dec{+27}{57}{33.37}\\
NGC 4921 & \RA{13}{01}{26.12} & \Dec{+27}{53}{09.59}\\
\enddata
\tablecomments{The cluster core is considered both as a point source and a modestly extended source. Three central galaxies are also considered in point-source searches. The choice is based on evidence of an excess of non-thermal X-ray emission \citep{article:Neumann_etal:2003} at the location of these galaxies.}
\label{table:roi}
\end{deluxetable}

\begin{deluxetable}{lccccccccc}
\tablecolumns{10}
\tablewidth{0pc}
\tablecaption{Flux upper limits for different regions of interest in the Coma cluster of galaxies and surroundings.}
\rotate
\tablehead{
\colhead{Source} & \colhead{RoI\tablenotemark{a} [deg]} & \colhead{$N_{S}$} & \colhead{S\tablenotemark{b} [$\sigma$]} &
\multicolumn{6}{c}{Flux U.L.\tablenotemark{c}} \\
\colhead{} & \colhead{} & \colhead{} & \colhead{} & \multicolumn{2}{c}{$\alpha=2.1$} &
\multicolumn{2}{c}{$\alpha=2.3$} &
\multicolumn{2}{c}{$\alpha=2.5$}
}
\startdata
Core & 0 & 17 & 0.84 & 2.59 & (0.78\%) & 2.78 & (0.83\%) & 2.97 & (0.89\%) \\
& 0.2 & -41 & -1.0 & 1.96 & (0.59\%)  & 2.09 & (0.63\%) & 2.21 & (0.66\%) \\
& 0.4 & -26 & -0.30 & 4.44 & (1.3\%)  & 4.74 & (1.4\%) & 5.02 & (1.5\%)\\
NGC 4889 & 0 & 3 & 0.14 & - & - & 1.85 & (0.55\%)  & - & - \\
NGC 4874 & 0 & -14 & -0.71 & - & - &  1.51 & (0.45\%)  & - & - \\
NGC 4921 & 0 & -4 & -0.23 & - & - &  2.41 & (0.72\%)  & - & -
\enddata
\tablenotetext{a}{Intrinsic source radius (zero means point source), which is convolved with the analysis PSF.}
\tablenotetext{b}{Statistical significance calculated according to \citet{article:LiMa:1983}.}
\tablenotetext{c}{99\% confidence level upper limit in units of $10^{-8}$ ph. m$^{-2}$ s$^{-1}$ calculated according to \citet{article:Rolke_etal:2005} above an energy threshold of 220 GeV, with corresponding fluxes in percent of the steady Crab Nebula flux in parenthesis.}
\label{table:results}
\end{deluxetable}

\begin{deluxetable}{cccccccccccc}
\tablecolumns{12}
\tablewidth{0pc}
\tablecaption{Constraints on relative CR pressure in the Coma cluster core for different spatial extensions.} 
\rotate
\tabletypesize{\small}
\tablehead{
\colhead{RoI\tablenotemark{a} [deg]} 
    & \colhead{$F_\rmn{sim}(E)$\tablenotemark{b}} & \colhead{$F_\rmn{sim}(>E)$\tablenotemark{c}} 
    & \multicolumn{3}{c}{analytic model: $X_{\CR,\,\rmn{max}}~[\%]$\tablenotemark{d} } 
    & \multicolumn{3}{c}{$F_{\gamma,\,\rmn{min}}(>E)$\tablenotemark{e}} 
    & \multicolumn{3}{c}{$X_{\CR,\,\rmn{min}}~[\%]$\tablenotemark{f} } \\
& & & \colhead{$\alpha=2.1$} & \colhead{$\alpha=2.3$} & \colhead{$\alpha=2.5$}
    & \colhead{$\alpha=2.1$} & \colhead{$\alpha=2.3$} & \colhead{$\alpha=2.5$}
    & \colhead{$\alpha=2.1$} & \colhead{$\alpha=2.3$} & \colhead{$\alpha=2.5$}
}
\startdata
0 & -     & -    & 10   & 23 & 97 & 1.6 & 0.7 & 0.3 & 0.067 & 0.061 & 0.11 \\
0.2  & 0.80  & 2.9  & 4.8  & 10 & 43 & 3.1 & 1.4 & 0.6 & 0.078 & 0.072 & 0.13 \\
0.4  & 1.21  & 4.4  & 6.7  & 15 & 62 & 6.3 & 2.8 & 1.3 & 0.098 & 0.090 & 0.16
\enddata
\tablenotetext{a}{Intrinsic source radius (zero means point source), which is convolved with the analysis PSF.}
\tablenotetext{b}{Differential gamma-ray flux $E^{2}dF/dE$ at $E=220$ GeV in units of $10^{-6}$ GeV m$^{-2}$ s$^{-1}$ in the simulation-based analytic model by \citet{article:PinzkePfrommer:2010}. }
\tablenotetext{c}{Integrated gamma-ray flux above $E=220~\rmn{GeV}$ in units of $10^{-9}$ ph. m$^{-2}$ s$^{-1}$; simulation-based analytical model by \citet{article:PinzkePfrommer:2010}.} 
\tablenotetext{d}{Constraint on the relative CR pressure, $X_{\CR} = \expval{P_\CR}/ \expval{P_\rmn{th}}$, which was assumed to be constant through the cluster and calculated according to \citet{article:PfrommerEnsslin:2004b}.}
\tablenotetext{e}{Minimum gamma-ray flux in units of $10^{-10}$ ph. m$^{-2}$ s$^{-1}$ of the hadronic model described in the text.}
\tablenotetext{f}{Minimum relative CR pressure, $X_{\CR,\,\rmn{min}} = F_{\rmn{min}}/F$, for energies $E>220~\rmn{GeV}$ in the hadronic model described in the text.}
\label{table:constraints}
\end{deluxetable}

\begin{deluxetable}{cccc}
\tablecolumns{4}
\tablewidth{0pc}
\tablecaption{Constraints on magnetic fields in the hadronic model of the radio halo of Coma cluster.}
\tablehead{
\colhead{$\alpha_{B}$} &
\multicolumn{3}{c}{Minimum magnetic field, $B_0~[\mu\rmn{G}]$}\\
\colhead{} & \colhead{$\alpha=2.1$} & \colhead{$\alpha=2.3$} & \colhead{$\alpha=2.5$}
}
\startdata
0.3 & 0.61 & 0.51 & 0.43 \\
0.5 & 0.82 & 0.68 & 0.58 \\
0.7 & 1.19 & 0.98 & 0.83 
\enddata
\tablecomments{The parameters of the magnetic field are the magnetic decline, $\alpha_B$, and the central field strength, $B_{0}$, which are defined by $B(r) = B_{0}\,[n_\rmn{e}(r)/n_{\rmn{e}}(0)]^{\alpha_B}$.}
\label{table:Bmin}
\end{deluxetable}

\begin{deluxetable}{lcccccc}
\tablecolumns{7}
\tablewidth{0pc}
\tablecaption{\emph{Fermi}-LAT upper limits for the Coma cluster core.}
\tablehead{
\colhead{Spatial model} & \multicolumn{3}{c}{S[$\sigma$]\tablenotemark{a}} & \multicolumn{3}{c}{Flux U.L.\tablenotemark{b}} \\
\colhead{} & \colhead{1-3 GeV} & \colhead{3-10 GeV} & \colhead{10-30 GeV} & 
\colhead{1-3 GeV} & \colhead{3-10 GeV} & \colhead{10-30 GeV} \\
\hline
\multicolumn{7}{c}{Spectral index $\alpha=2.1$ (fixed)}
}
\startdata
Point source & 0.30 & 0 & 1.3 & 1.73 & 4.87 & 0.627 \\
Disk, $r=0.2^{\circ}$ & 0.65 & 0.22 & 2.0 & 2.03 & 0.691 & 0.783 \\
Disk, $r=0.4^{\circ}$ & 0.77 & 0.18 & 2.3 & 2.39 & 0.757 & 0.847 \\
\cutinhead{Spectral index $\alpha=2.5$ (fixed)}
Point source & 0.33 & 0 & 1.42 & 1.81 & 0.526 & 0.635 \\
Disk, $r=0.2^{\circ}$ & 0.66 & 0.35 & 2.1 & 2.12 & 0.748 & 0.785 \\
Disk, $r=0.4^{\circ}$ & 0.80 & 0.36 & 2.4 & 2.50 & 0.833 & 0.849 \\
\enddata
\tablenotetext{a}{Statistical significance.}
\tablenotetext{b}{95\% confidence level upper limit in units of $10^{-10}$ ph. cm$^{-2}$ s$^{-1}$.}
\label{table:fermi}
\end{deluxetable}

%\begin{deluxetable}{lccc}
%\tablecolumns{4}
%\tablewidth{0pc}
%\tablecaption{Limits on the DM annihilation cross section $\expval{\sigma v}$.}
%\tablehead{
%\colhead{Channel} & RoI [deg] & $m_{\chi}$ [GeV] & $\expval{\sigma v}$ [cm$^{3}$ s$^{-1}$] \\
%}
%\startdata
%W$^{+}$W$^{-}$ & & & \\
%$b\bar{b}$ & & & \\
%$\tau^{+}\tau^{-}$ & & & \\
%\enddata
%\tablecomments{Limits are for different integration regions and DM particle mass $m_{\chi}$ derived from %the VERITAS gamma-ray flux upper limits presented in this work.}
%\end{deluxetable}

>>>>

A third mechanism for gamma-ray production in a galaxy cluster is self-annihilation of a DM particle, e.g. a weakly interacting massive particle (WIMP). As already mentioned, about 80\% of the cluster mass is in the form of dark matter. Recent works \citep{article:EvansFerrerSarkar:2004, article:BergstromHooper:2006} have shown that the gamma-ray emissivity from DM annihilations is expected to be proportional to the DM density along the line of sight, making clusters of galaxies potentially good candidates for DM searches, despite their large distances.

@@

\begin{deluxetable}{lcc}
\tablecolumns{3}
\tablewidth{0pc}
\tablecaption{Regions of interest in the Coma cluster field of view}
\tablehead{
\colhead{Source} &
\colhead{RA} &
\colhead{Dec}
}
\startdata
Core & \RA{12}{59}{48.7} & \Dec{+27}{58}{50.0}\\
NGC 4889 & \RA{13}{00}{08.13} & \Dec{+27}{58}{37.03}\\
NGC 4874 & \RA{12}{59}{35.71} & \Dec{+27}{57}{33.37}\\
NGC 4921 & \RA{13}{01}{26.12} & \Dec{+27}{53}{09.59}\\
\enddata
\tablecomments{The cluster core is considered both as a point source and a modestly extended source. Three central galaxies are also considered in point-source searches. The choice is based on evidence of an excess of non-thermal X-ray emission \citep{article:Neumann_etal:2003} at the location of these galaxies.}
\label{table:roi}
\end{deluxetable}

\begin{deluxetable}{lccccccccc}
\tablecolumns{10}
\tablewidth{0pc}
\tablecaption{VERITAS flux upper limits for different regions of interest in the Coma cluster of galaxies and surroundings}
\rotate
\tablehead{
\colhead{Source} & \colhead{RoI\tablenotemark{a} [deg]} & \colhead{$N_{S}$} & \colhead{S\tablenotemark{b} [$\sigma$]} &
\multicolumn{6}{c}{Flux U.L.\tablenotemark{c}} \\
\colhead{} & \colhead{} & \colhead{} & \colhead{} & \multicolumn{2}{c}{$\alpha=2.1$} &
\multicolumn{2}{c}{$\alpha=2.3$} &
\multicolumn{2}{c}{$\alpha=2.5$}
}
\startdata
Core & 0 & 17 & 0.84 & 2.59 & (0.78\%) & 2.78 & (0.83\%) & 2.97 & (0.89\%) \\
& 0.2 & -41 & -1.0 & 1.96 & (0.59\%)  & 2.09 & (0.63\%) & 2.21 & (0.66\%) \\
& 0.4 & -26 & -0.30 & 4.44 & (1.3\%)  & 4.74 & (1.4\%) & 5.02 & (1.5\%)\\
NGC 4889 & 0 & 3 & 0.14 & - & - & 1.85 & (0.55\%)  & - & - \\
NGC 4874 & 0 & -14 & -0.71 & - & - &  1.51 & (0.45\%)  & - & - \\
NGC 4921 & 0 & -4 & -0.23 & - & - &  2.41 & (0.72\%)  & - & -
\enddata
\tablenotetext{a}{Intrinsic source radius (zero means point source), which is convolved with the analysis PSF.}
\tablenotetext{b}{Statistical significance calculated according to \citet{article:LiMa:1983}.}
\tablenotetext{c}{99\% confidence level upper limit in units of $10^{-8}$ ph. m$^{-2}$ s$^{-1}$ calculated according to \citet{article:Rolke_etal:2005} above an energy threshold of 220 GeV, with corresponding fluxes in percent of the steady Crab Nebula flux in parenthesis.}
\label{table:results}
\end{deluxetable}

\begin{deluxetable}{cccccccccccc}
\tablecolumns{12}
\tablewidth{0pc}
\tablecaption{Constraints on relative CR pressure in the Coma cluster core for different spatial extensions} \rotate
\tabletypesize{\small}
\tablehead{
\colhead{RoI\tablenotemark{a} [deg]} 
    & \colhead{$F_\rmn{sim}(E)$\tablenotemark{b}} & \colhead{$F_\rmn{sim}(>E)$\tablenotemark{c}} 
    & \multicolumn{3}{c}{analytic model: $X_{\CR,\,\rmn{max}}~[\%]$\tablenotemark{d} } 
    & \multicolumn{3}{c}{$F_{\gamma,\,\rmn{min}}(>E)$\tablenotemark{e}} 
    & \multicolumn{3}{c}{$X_{\CR,\,\rmn{min}}~[\%]$\tablenotemark{f} } \\
& & & \colhead{$\alpha=2.1$} & \colhead{$\alpha=2.3$} & \colhead{$\alpha=2.5$}
    & \colhead{$\alpha=2.1$} & \colhead{$\alpha=2.3$} & \colhead{$\alpha=2.5$}
    & \colhead{$\alpha=2.1$} & \colhead{$\alpha=2.3$} & \colhead{$\alpha=2.5$}
}
\startdata
0 & -     & -    & 10   & 23 & 97 & 1.6 & 0.7 & 0.3 & 0.067 & 0.061 & 0.11 \\
0.2  & 0.80  & 2.9  & 4.8  & 10 & 43 & 3.1 & 1.4 & 0.6 & 0.078 & 0.072 & 0.13 \\
0.4  & 1.21  & 4.4  & 6.7  & 15 & 62 & 6.3 & 2.8 & 1.3 & 0.098 & 0.090 & 0.16
\enddata
\tablenotetext{a}{Intrinsic source radius (zero means point source), which is convolved with the analysis PSF.}
\tablenotetext{b}{Differential gamma-ray flux $E^{2}dF/dE$ at $E=220$ GeV in units of $10^{-6}$ GeV m$^{-2}$ s$^{-1}$ in the simulation-based analytic model by \citet{article:PinzkePfrommer:2010}. }
\tablenotetext{c}{Integrated gamma-ray flux above $E=220~\rmn{GeV}$ in units of $10^{-9}$ ph. m$^{-2}$ s$^{-1}$; simulation-based analytical model by \citet{article:PinzkePfrommer:2010}.} 
\tablenotetext{d}{Constraint on the relative CR pressure, $X_{\CR} = \expval{P_\CR}/ \expval{P_\rmn{th}}$, which was assumed to be constant through the cluster and calculated according to \citet{article:PfrommerEnsslin:2004b}.}
\tablenotetext{e}{Minimum gamma-ray flux in units of $10^{-10}$ ph. m$^{-2}$ s$^{-1}$ of the hadronic model described in the text.}
\tablenotetext{f}{Minimum relative CR pressure, $X_{\CR,\,\rmn{min}} = F_{\rmn{min}}/F$, for energies $E>220~\rmn{GeV}$ in the hadronic model described in the text.}
\label{table:constraints}
\end{deluxetable}

\begin{deluxetable}{cccc}
\tablecolumns{4}
\tablewidth{0pc}
\tablecaption{Constraints on magnetic fields in the hadronic model of the radio halo of Coma cluster}
\tablehead{
\colhead{$\alpha_{B}$} &
\multicolumn{3}{c}{Minimum magnetic field, $B_0~[\mu\rmn{G}]$}\\
\colhead{} & \colhead{$\alpha=2.1$} & \colhead{$\alpha=2.3$} & \colhead{$\alpha=2.5$}
}
\startdata
0.3 & 0.61 & 0.51 & 0.43 \\
0.5 & 0.82 & 0.68 & 0.58 \\
0.7 & 1.19 & 0.98 & 0.83 
\enddata
\tablecomments{The parameters of the magnetic field are the magnetic decline, $\alpha_B$, and the central field strength, $B_{0}$, which are defined by $B(r) = B_{0}\,[n_\rmn{e}(r)/n_{\rmn{e}}(0)]^{\alpha_B}$.}
\label{table:Bmin}
\end{deluxetable}

\begin{deluxetable}{lcccccc}
\tablecolumns{7}
\tablewidth{0pc}
\tablecaption{\emph{Fermi}-LAT upper limits for the Coma cluster core}
\tablehead{
\colhead{Spatial model} & \multicolumn{3}{c}{S[$\sigma$]\tablenotemark{a}} & \multicolumn{3}{c}{Flux U.L.\tablenotemark{b}} \\
\colhead{} & \colhead{1-3 GeV} & \colhead{3-10 GeV} & \colhead{10-30 GeV} & 
\colhead{1-3 GeV} & \colhead{3-10 GeV} & \colhead{10-30 GeV} \\
\hline
\multicolumn{7}{c}{Spectral index $\alpha=2.1$ (fixed)}
}
\startdata
Point source & 0.30 & 0 & 1.3 & 1.73 & 4.87 & 0.627 \\
Disk, $r=0.2^{\circ}$ & 0.65 & 0.22 & 2.0 & 2.03 & 0.691 & 0.783 \\
Disk, $r=0.4^{\circ}$ & 0.77 & 0.18 & 2.3 & 2.39 & 0.757 & 0.847 \\
\cutinhead{Spectral index $\alpha=2.5$ (fixed)}
Point source & 0.33 & 0 & 1.42 & 1.81 & 0.526 & 0.635 \\
Disk, $r=0.2^{\circ}$ & 0.66 & 0.35 & 2.1 & 2.12 & 0.748 & 0.785 \\
Disk, $r=0.4^{\circ}$ & 0.80 & 0.36 & 2.4 & 2.50 & 0.833 & 0.849 \\
\enddata
\tablenotetext{a}{Statistical significance.}
\tablenotetext{b}{95\% confidence level upper limit in units of $10^{-10}$ ph. cm$^{-2}$ s$^{-1}$.}
\label{table:fermi}
\end{deluxetable}

\begin{deluxetable}{ccc}
\tablecolumns{3}
\tablewidth{0pc}
\tablecaption{Astrophysical factors}
\tablehead{
\colhead{RoI [deg]} & $\expval{J}_{\rmn{signal}}$ [GeV$^{2}$ cm$^{-5}$ sr] & $\alpha\expval{J}_{\rmn{bkg}}$ [GeV$^{2}$ cm$^{-5}$ sr]
}
\startdata
0 & $5.7\times 10^{16}$ & $1.3\times 10^{14}$ (negligible) \\
0.2 & $8.1\times 10^{16}$ & $4.4\times 10^{14}$ ($<0.01\expval{J}_{\rmn{signal}}$, negligible) \\
0.4 & $9.4\times 10^{16}$ & $1.3\times 10^{15}$ ($\simeq0.01\expval{J}_{\rmn{signal}}$, negligible)
\enddata
\tablecomments{$\expval{J}_{\rmn{bkg}}$ is the astrophysical factor calculated for the background region (ring method) and is used to estimate the level of gamma-ray contamination from DM.}
\end{deluxetable}

\begin{deluxetable}{lccc}
\tablecolumns{4}
\tablewidth{0pc}
\tablecaption{Limits on the DM annihilation cross section $\expval{\sigma v}$.}
\tablehead{
\colhead{Channel} & RoI [deg] & $m_{\chi}$ [GeV] & $\expval{\sigma v}$ [cm$^{3}$ s$^{-1}$]
}
\startdata
W$^{+}$W$^{-}$ & 0 & 2000 & $1.1\times 10^{-20}$ \\
& 0.2 & 1900 & $4.3\times 10^{-21}$ \\
& 0.4 & 1900 & $8.4\times 10^{-21}$ \\
$b\bar{b}$ & 0 & 3500 & $1.2\times 10^{-20}$ \\
& 0.2 & 3400 & $4.4\times 10^{-21}$ \\
& 0.4 & 3500 & $8.7\times 10^{-21}$ \\
$\tau^{+}\tau^{-}$ & 0 & 670 & $2.4\times 10^{-21}$ \\
& 0.2 & 650 & $9.1\times 10^{-22}$ \\
& 0.4 & 660 & $1.8\times 10^{-21}$ \\
\enddata
\tablecomments{Limits are for different integration regions and DM particle mass $m_{\chi}$ derived from the VERITAS gamma-ray flux upper limits presented in this work.}
\end{deluxetable}
